### 2016-04-22 manuscript-body.tex
Drive home points about structural and syntactic similarity
Add example statements in javascript

That Bonar and Soloway would even consider developing new constructs or
languages is a noteworthy distinction. Rather than assume wholesale that
students with ``bugs'' had wrong knowledge, the authors instead suppose
cognitive bugs have plausible origins worth designing around. Moreover,
they treat students' divergence from canon as \emph{an opportunity for
research to learn from students}. Precisely because students saw
functional differences (cf., diSessa, 1986) in semantically-equivalent
PASCAL statements, Bonar and Soloway (1983) reflect that perhaps
programming languages should be more expressive:

@@

Let's begin with two motivating examples. My aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. In Table 2 below, we imagine five
different ways of writing a programming statement. Each of the five ways
is semantically equivalent to the others (or near enough for explanatory
purposes). And, in each example we add a layer of specificity to the
syntax. we also propose a corresponding interpretation of how we might
apply interpret and think about a given statement.\footnote{we assume for
  simplicity's sake that all variables and functions in my code samples
  evaluate to floating-point numbers that can be combined under
  addition. we also assume, crucially, that all statements are in the
  same hypothetical language and that the variable names connote nothing
  to the computational interpreter.}

\protect\hypertarget{ux5fToc252445956}{}{}Table -- Five different
interpretations of semantically equivalent programming statements in the
same language

\begin{center}
  \begin{tabular}{ | l | l |}
    1 & \texttt{x\ =\ x\ +\ d} \\
    2 & \texttt{x\ =\ x\ +\ update} \\
    3 & \texttt{x\_position\ =\ x\_position\ +\ update} \\
    4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)} \\
    5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)} \\
  \end{tabular}
\end{center}

we stress that the third column is about \emph{how we might think about}
each programming statement\emph{.} By no means am we making normative
claims about how one \emph{ought} to think about it, or whether the
statement actually does what its names might suggest it does. Rather,
``How we might think about it'' reflects the kind of local meaning or
interpretation we might attach to such a statement when we work with it,
given my understanding of its role and context.

Consider specifically statements 1 and 4. Statement 1 could very well be
an incrementer in some kind of iterative code. If we had to debug code
that employs statement 1, what we might do is exploit my knowledge of
what d is (does it hold a constant value? Is it 1?) and try inserting
intermediate print statements into the iterative code. Doing so, we can
inspect textually how values change with each iteration. In statement
4's case, it could very well be that the position-update code animates
images on a screen. If there's a problem with that code, one of the
easiest ways we might notice is that the acceleration seems off in the
graphics. Consequently, my debugging might call upon knowledge we have
from kinematics. we might try inserting code that draws a dot at the
object's on-screen position with each iteration, leaving a trail we can
visually inspect. we could then look at the path of the object's
trajectory and the spacing patterns between successive dots as a
first-pass test of whether my code achieves the motion we want.

To be clear, it's not just that graphics might improve my efficiency in
debugging a statement. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets me
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in my code. If we view the assignment statement as saying
something about the motion of an object (cf., Hammer, 1994, p. 165), a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if we focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what we know about physics in order to
reason about the code.

My second example concerns statements that are syntactically-similar,
rather than semantically-similar. we proposed that the statements in
Table 2 all came from the same language. But, another phenomenon comes
into play when different languages use the same symbology for
conceptually different operations. Table 3 shows examples of what look
like semantically-equivalent operations, but in fact are not.

\protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
syntactically-similar statements with very different semantics

@@

The catch here is that the plus operator takes on different roles in
different languages because of how those languages define its use.
Statement 1 increments a number in C; Statement 2 appends the letters
``ly'' to a string; Statement 3 adds a layer of points to a statistical
graphics plot. These different kinds of operations become even more
apparent and consequential when, for example, such statements are
repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
  operator can be written as a directional arrow. The symbol
  \textless{}- (``less-than, hyphen'') indicates the value on the right
  side of the symbol is being assigned to the variable on the left side
  of the symbol. As used, the symbol itself is semantically equivalent
  to having written an equals sign (``=``).} I'm using multiple
reassignment statements to compose a statistical graphics plot. The
layered creation of a plot invites a very different kind of conceptual
interpretation than, say, repeatedly accumulating numbers into a running
sum:

p \textless{}- InitializeGgplot\_1w()

p \textless{}- p + GrandMeanLine(owp)

p \textless{}- p + GrandMeanPoint(owp)

p \textless{}- p + ScaleX\_1w(owp)

p \textless{}- p + ScaleY\_1w(owp)

p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)

One obvious reason for thinking about this code with a different
interpretive frame is that numeric addition is commutative; composing a
plot is not necessarily commutative.\footnote{To convince yourself that
  plot composition isn't commutative, imagine a scale function that
  squares up the aspect ratio of a plot and another scale function that
  sets the aspect ratio to 1.5:1. Applying the square function last
  produces a square plot; applying it first produces a rectangular plot.
  String concatenation is also not necessarily commutative. ``cat'' +
  ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
  ``dogcat.''} So, despite the syntactic similarity, reassignments that
compose a plot using reassignment (as above) does not obey the same
rules as reassignments for a running total. But, even within this code
block, statements that look alike perform operations of a different
nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
visual layer to a plot, others modify features of the plot (e.g, ``+
Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
the numeric range of the data).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Programming can be helped by applying conceptual models to code,
  particularly when relevant domain-knowledge structures can
  advantageously transform a problem (example 1)
\item
  But, conceptual models don't work all the time for all statements.
  Because languages are designed differently, the same syntax can
  actually correspond to very different operations in code (example 2).
  And, that's true both within and across languages.
\item
  Consequently, it makes less sense to treat conceptual models as right
  or wrong, and more sense to treat them as differentially advantageous
  for thinking about what a piece of code does. (Thinking a ``+''
  implies numerical addition isn't \emph{globally wrong} in JavaScript,
  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
\item
  It seems plausible that successful programmers, when reasoning about
  or writing code, are able to dynamically access or deploy conceptual
  models that are advantageous given the context (language, syntax,
  surrounding code). Certainly such a supposition is in line with work
  suggesting students have resources for thinking conceptually and
  dynamically about how mathematics models real-world situations (Izsák,
  2004; Sherin, 2001)
\item
  To model how programmers think with conceptual models, a suitable
  framework should be able to account for the dynamic, context-sensitive
  deployment of conceptual knowledge.
\item
  To model how programmers develop expertise, a suitable framework
  should be able to describe higher-order phenomena. Such phenomena
  include explaining how programmers come to have conceptual models or
  generate new ones, why they decide to deploy them, and how programmers
  consider which conceptual model (i.e., which way to think about code)
  is appropriate.
\end{enumerate}

>>>>

That Bonar and Soloway would even consider developing new constructs or
languages is a noteworthy distinction. Rather than assume wholesale that
students with ``bugs'' had wrong knowledge, the authors instead suppose
cognitive bugs have plausible origins worth designing around. Moreover,
they treat students' divergence from canon as \emph{an opportunity for
research to learn from students}. Precisely because students saw
functional differences in semantically-equivalent
PASCAL statements, Bonar and Soloway \cite{bonar_uncovering_1983} reflect that perhaps
programming languages should be more expressive:

@@

Let's begin with two motivating examples. Our aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. Below are JavaScript statements that, as programmers, we have actually written in code as part of our careers. Each statement
adheres to the structural similarity in Bonar and Soloway's PASCAL example \cite{bonar_uncovering_1983} discussed earlier. But, the comment above each statement reflects the \emph{functional understandings} \cite{disessa_models_1986} we made use of to write, interpret, and debug the code we were working on.

\begin{Shaded}
\begin{Highlighting}[numbers=left,,]
\CommentTok{// Increment a counter}
\NormalTok{x }\OperatorTok{=} \NormalTok{x }\OperatorTok{+} \DecValTok{1}

\CommentTok{// Advance a simulation forward}
\NormalTok{t }\OperatorTok{=} \NormalTok{t }\OperatorTok{+} \NormalTok{dt}

\CommentTok{// Move an object under uniform velocity}
\NormalTok{x_position }\OperatorTok{=} \NormalTok{x_position }\OperatorTok{+} \NormalTok{displacement}

\CommentTok{// Accelerate an object along the x-dimension}
\NormalTok{x_position }\OperatorTok{=} \NormalTok{x_position }\OperatorTok{+} \AttributeTok{displacement}\NormalTok{(t)}

\CommentTok{// Jitter a plot point using a stochastic function}
\NormalTok{y_position }\OperatorTok{=} \NormalTok{y_position }\OperatorTok{+} \AttributeTok{random_noise}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

We stress the comments reflect \emph{how one might think about}
each programming statement. By no means are we making normative
claims about how one \emph{ought} to think about it, or whether the
statement actually does what its variable names might suggest it does. Rather,
``How we might think about it'' reflects the kind of local meaning or
interpretation we might attach to such a statement when we work with it,
given our understanding of its role and context.

Consider specifically lines 2 and 11. Line 2 could very well be
an incrementer in some kind of iterative code. If we had to debug loop code
that employs line 2, what we might do is try inserting
intermediate print statements to see the value of x. We might also call on our knowledge that it's being incremented by 1 to check whether the difference between any two successive printouts is 1; if it isn't, we immediately know something's wrong. In line
11's case, it could very well be that line 11's function is animating
images on a screen. If there's a problem with that code, one of the
easiest ways we might notice is that the acceleration seems off in the
animation. Consequently, debugging might call upon knowledge we have
from kinematics. We might try inserting code that draws a dot at the
object's on-screen position with each iteration, leaving a trail of dots we can
visually inspect. We could then look at the path of the object's
trajectory and the spacing patterns between successive dots as a
first-pass test of whether the code achieves the motion we want. (If, for example, we're attempting a cubic easing animation, but the dots are all evenly spaced, then we might look first at the \texttt{displacement(t)}) function and what it's returning.

To be clear, it's not just that graphics might improve our efficiency in
debugging a statement like line 11. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets us
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in code. If we view the assignment statement as proposing
something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if we focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what we know about physics in order to
reason about the code.

Our second example concerns statements that use the same symbol template for fundamentally different kinds of computation. In our JavaScript example statements, all code came from the same language. But, another phenomenon comes
into play \emph{across} different languages when use the same symbology (written patterns) to stand for
conceptually different operations. The statements below are examples of what look
like semantically-equivalent operations, but in fact are not.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{// C - Increments the value of i by 1}
\NormalTok{i = i + }\DecValTok{1}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{// JavaScript - Appends "ly" to word.toString()}
\NormalTok{word }\OperatorTok{=} \NormalTok{word }\OperatorTok{+} \StringTok{"ly"}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# R/ggplot2 - Composes a layer of points onto a plot}
\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

The catch here is that the plus operator takes on different roles in
different languages because of how those languages define its use.
Statement 1 increments a number in C; Statement 2 appends the letters
``ly'' to a string; Statement 3 composes a layer of points onto a statistical
graphics plot. These different kinds of operations become even more
apparent and consequential when, for example, such statements are
repeated in the same language within the same file. In the R/ggplot2 code below, the author uses multiple
reassignment statements to compose a statistical graphics plot.

\begin{Shaded}
\begin{Highlighting}[numbers=left,,]
\NormalTok{p <-}\StringTok{ }\KeywordTok{InitializeGgplot_1w}\NormalTok{()}
\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanLine}\NormalTok{(owp)}
\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanPoint}\NormalTok{(owp)}
\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleX_1w}\NormalTok{(owp)}
\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleY_1w}\NormalTok{(owp)}
\end{Highlighting}
\end{Shaded}

Despite the syntactic similarity, The
layered creation of a plot invites a very different kind of conceptual
interpretation than, say, repeatedly accumulating numbers into a running
sum. One obvious reason for thinking about this code with a different
interpretive frame than incrementation is that numeric addition is commutative; composing a
plot is not necessarily commutative. So, despite the syntactic similarity, reassignments that
compose a plot using reassignment (as above) do not obey the same
rules as reassignments for a running total. But, it turns out the statements above don't obey the same rules of string concatenation either. Within this code
block, statements that look alike perform operations of a different
nature. While some expressions (lines 2 and 3) compose
visual layers onto a plot, others (lines 4 and 5) modify features of the plot, like the x and y scales.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Programming can be helped by applying conceptual models to code,
  particularly when relevant domain-knowledge structures can
  advantageously transform a problem (example 1)
\item
  But, conceptual models don't work all the time for all statements.
  Because languages are designed differently, the same syntax can
  actually correspond to very different operations in code (example 2).
  And, that's true both within and across languages.
\item
  Consequently, it makes less sense to treat conceptual models as right
  or wrong, and more sense to treat them as differentially advantageous
  for thinking about what a piece of code does. (Thinking a ``+''
  implies numerical addition isn't \emph{globally wrong} in JavaScript,
  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
\item
  It seems plausible that successful programmers, when reasoning about
  or writing code, are able to dynamically access or deploy conceptual
  models that are advantageous given the context (language, syntax,
  surrounding code). Certainly prior research demonstrates novices and experts both have resources for creating, evaluating, and adapting their own conceptual metaphors to suit the context of the problem \cite{hammer_static_2011,izsak_we_2003,izsak_students_2004}
\item
  To model how programmers think with conceptual models, a suitable
  framework should be able to account for the dynamic, context-sensitive
  deployment of conceptual knowledge.
\item
  To model how programmers develop expertise, a suitable framework
  should be able to describe higher-order phenomena. Such phenomena
  include explaining how programmers come to have conceptual models or
  generate new ones, why they decide to deploy them, and how programmers
  consider which conceptual model (i.e., which way to think about code)
  is appropriate.
\end{enumerate}
